---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 7,
  fig.height = 3.2,
  dpi = 120,
  out.width = "100%"
)
```

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

# ebw

## Overview of `ebw`

The `ebw` package estimates covariate balancing weights for causal inference by
energy balancing, which reweights the covariate distributions so that the treatment
groups come to resemble a common target distribution. Because energy balancing acts
on the whole distribution instead of a chosen set of moments, and a distribution
balanced in this sense is balanced in every moment at once, the resulting weights
tend to be robust to the way the outcome depends on the covariates and can work
quite well across a range of settings.

A single function, `energy_balance()`, covers three kinds of treatment. For a
binary treatment it returns the energy balancing weights of Huling and Mak (2024);
for a treatment with several categories it balances each group to the pooled
sample; and for a continuous treatment it returns the independence weights of
Huling, Greifer, and Chen (2024), which reweight the sample toward independence
between the treatment and the covariates. An improved variant additionally
balances the treatment groups to one another.

Every fit records the ingredients of its dual representation, which expresses each
weight as an explicit function of the covariates. Evaluating that function at a new
covariate value gives the weight a unit outside the fitted sample would receive, so
`predict()` returns out-of-sample weights without refitting. The package brings
binary, multi-category, and continuous treatments, together with out-of-sample
prediction, under one interface, and is intended to grow into a single home for
this family of methods, which have until now been split across separate packages.

## Installing the `ebw` package

Install the development version from GitHub:

``` r
# install.packages("remotes")
remotes::install_github("jaredhuling/ebw")
```

The quadratic programs are solved with `osqp`, and the diagnostic plots use
`ggplot2`; both are installed automatically.

## Quick usage overview

### A binary treatment

Supply a covariate matrix and a treatment vector, or a formula and a data frame.
The two interfaces produce the same weights.

```{r binary}
library(ebw)

set.seed(1)
n <- 300; p <- 5
X <- matrix(rnorm(n * p), n, p)
a <- rbinom(n, 1, plogis(X[, 1] - X[, 2]))

fit <- energy_balance(x = X, treatment = a, estimand = "ATE")
fit
```

The covariates are standardized by default before the kernel is formed, following
the scaling conventions that make weighting methods behave well when the covariates
are measured on different scales. The `scaling` argument controls this behavior.

### The improved estimator

Setting `improved = TRUE` adds a term that matches the weighted treatment groups to
one another as well as to the target, which reduces the residual imbalance between
the groups.

```{r improved}
fit_improved <- energy_balance(x = X, treatment = a, improved = TRUE)
```

### A multi-category treatment

A treatment that is a factor, or that takes a small number of distinct values, is
balanced group by group to the pooled sample. Here the three groups are assigned
in a way that depends on the covariates, so that the groups differ before
weighting and there is imbalance to remove.

```{r multi}
dat <- as.data.frame(X)
lp <- drop(X %*% c(1, -1, 0, 0, 0))
prob <- exp(cbind(-lp, 0, lp)); prob <- prob / rowSums(prob)
dat$A <- factor(
  c("low", "medium", "high")[apply(prob, 1, function(pr) sample(1:3, 1, prob = pr))],
  levels = c("low", "medium", "high"))

fit_multi <- energy_balance(A ~ ., data = dat)
```

`balance()` reports the energy distance of each group to the pooled sample, which
the weights reduce substantially, along with the standardized mean differences and
the between-group energy distances.

```{r multi-balance}
balance(fit_multi)
```

### A continuous treatment

A numeric treatment with many distinct values is handled by independence weights,
which drive the weighted distance covariance between the treatment and the
covariates toward zero while preserving the marginal distributions.

```{r continuous}
A_cont <- drop(X %*% rep(0.2, p)) + rnorm(n)
fit_cont <- energy_balance(x = X, treatment = A_cont)
```

### Predicting weights out of sample

Because the fit carries its dual representation, the weight a new unit would
receive can be predicted from its covariates (and, for a discrete or continuous
treatment, its treatment value) without refitting.

```{r predict}
Xnew <- matrix(rnorm(20 * p), 20, p)
anew <- rbinom(20, 1, 0.5)
w_new <- predict(fit, newdata = Xnew, treatment = anew)
summary(w_new)
```

### Assessing balance and inspecting the weights

`balance()` reports the covariate balance the weights achieve, in the fitted sample
or, when `newdata` is supplied, in a fresh sample under the predicted weights.

```{r balance}
balance(fit)
```

`plot()` draws a Love plot of the balance and, with `type = "weights"`, a summary
of the weight distribution.

```{r loveplot}
plot(fit)
```

```{r weightplot}
plot(fit, type = "weights")
```

The weights returned by `weights()` are a plain numeric vector, so they pass
directly to downstream tools such as `cobalt::bal.tab()` for further balance
assessment.

## Accessing help files for the main functions

``` r
?energy_balance       # fit weights for binary, multi-category, or continuous treatments
?predict.ebw          # out-of-sample weight prediction from the dual representation
?balance              # in-sample and out-of-sample balance diagnostics
?independence_weights # continuous-treatment interface compatible with independenceWeights
```

## References

Huling, J. D. and Mak, S. (2024). Energy balancing of covariate distributions.
*Journal of Causal Inference*, 12(1):20220029.

Huling, J. D., Greifer, N., and Chen, G. (2024). Independence weights for causal
inference with continuous treatments. *Journal of the American Statistical
Association*, 119(546):1657--1670.
