---
title: "Energy balancing and independence weights with ebw"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Energy balancing and independence weights with ebw}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
library(ebw)
set.seed(20260809)
```

The **ebw** package fits covariate balancing weights for causal inference with
either a binary treatment, through energy balancing weights, or a continuous
treatment, through independence weights. A single function,
`energy_balance()`, handles both. Every fit records the pieces of its dual
representation, so the weight a new unit would receive can be predicted out of
sample with `predict()`.

## Binary treatment: energy balancing

```{r binary-fit}
dat <- sim_confounded_data(n.obs = 300, n.vars = 6, AR.cor = 0.3)
fit_b <- energy_balance(x = dat$x, treatment = dat$trt, estimand = "ATE")
fit_b
```

The fitted weights are returned by `weights()` as a plain numeric vector, which
slots directly into downstream balance tools.

```{r binary-weights}
w <- weights(fit_b)
str(w)
```

### Out-of-sample prediction

The fit records the quantities that define the weight as an explicit function of
the covariates: the fitted weights, the kernel evaluations of a covariate value
against the training units, the fixed centering constants, the threshold
multiplier, and the ridge parameter. Evaluating that function at a new covariate
value gives the weight a genuinely new unit would receive, so `predict()` returns
out-of-sample weights without refitting.

```{r binary-predict}
new <- sim_confounded_data(n.obs = 50, n.vars = 6, AR.cor = 0.3)
w_new <- predict(fit_b, newdata = new$x)
summary(w_new)
```

### Balance diagnostics

`balance()` reports per-covariate standardized mean differences before and after
weighting, the energy distance to the target, and the effective sample size. It
works both on the training sample and, with `newdata`, out of sample.

```{r binary-balance}
balance(fit_b)
balance(fit_b, newdata = list(x = new$x, trt = new$trt))
```

A Love plot and a plot of the weight distribution are available through
`plot()`.

```{r binary-plot, fig.alt = "Love plot of covariate balance"}
plot(fit_b, type = "balance")
plot(fit_b, type = "weights")
```

### Interoperability with cobalt

Because `weights()` returns a plain numeric vector, the fitted weights pass
directly to `cobalt::bal.tab()`.

```{r cobalt, eval = requireNamespace("cobalt", quietly = TRUE)}
cobalt::bal.tab(dat$x, treat = dat$trt, weights = weights(fit_b))
```

## Continuous treatment: independence weights

For a continuous treatment, `independence_weights()` fits weights that drive the
weighted joint distribution of covariates and treatment toward independence.

```{r cont-fit}
cdat <- sim_continuous_data(n.obs = 300, n.vars = 4, alpha.norm2 = 0.5)
fit_c <- independence_weights(A = cdat$a, X = cdat$x)
fit_c
```

Prediction is again out of sample, but the balancing function now depends on
both the covariates and the treatment value, so both are required.

```{r cont-predict}
cnew <- sim_continuous_data(n.obs = 50, n.vars = 4, alpha.norm2 = 0.5)
w_cnew <- predict(fit_c, newdata = list(x = cnew$x, a = cnew$a))
summary(w_cnew)
```

The continuous balance report gives per-covariate distance correlations with the
treatment, the overall distance covariance, and the weighted versus unweighted
marginal means.

```{r cont-balance}
balance(fit_c)
```

### Moment constraints

Two optional moment constraints refine the continuous fit. `preserve_means`
forces the weighted mean of every covariate and of the treatment to equal the
unweighted sample mean, and `decorrelate_moments` forces the weighted
treatment-covariate covariances to zero. Both are solved with `osqp`, and the
out-of-sample predictor gains a dual term that is linear in the constraint
features, so prediction stays out of sample.

```{r cont-constrained}
fit_pm <- independence_weights(A = cdat$a, X = cdat$x, preserve_means = TRUE)
w_pm <- weights(fit_pm)

# the weighted covariate means now match the unweighted means
round(max(abs(colSums(w_pm * cdat$x) / sum(w_pm) - colMeans(cdat$x))), 12)

# and prediction remains available out of sample
head(predict(fit_pm, newdata = list(x = cnew$x, a = cnew$a)))
```
